Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-28 paper.tex
Updates

\subsection{Future applications: Cloud computing}
To explore the use of cloud computing services with VCF Zarr,
Genomics England provided access to data within their secure
cloud environment. The VCF Zarr files for Chromosome 2
described in Table~\ref{tab-genomics-england-data}
were hosted in a standard AWS S3 bucket.

Using this data we repeated the benchmarks above on the same subset
of 562,640 variants and 78,195 samples using the same
single-threaded sequential Zarr code.
Extracting the position array required 15s CPU, 23.8s elapsed.
The af-dist calculation required 2.70min CPU, 5.23min elapsed.
Creating the filter mask required 3.05min CPU, 11.15min elapsed.

While these benchmarks are comparable with results obtained on the
shared file-system in the Genomics England research environment,
the sequential single-threaded access patterns used do not map
well to modern cloud computing environments.
As each request to a cloud object store has a high latency, it
is important to have many requests running concurrently to achieve
good performance~\citep{durner2023exploiting}.
To explore the limits of performance we developed a simple prototype
Zarr implementation, designed to maximise concurrency using asynchrous
IO and multiple processes (see Methods for details).
Fig~\ref{fig-s3-throughput} shows the rate at which genotypes
can be decoded to RAM as well as rate at which we can compute
af-dist (using Numba to classify genotypes, as before)
as we increase the number of parallel processes on
an EC2 c5.24xlarge instance with 96 vCPUs (48 physical).
The maximum decompression rate of 47.4GiB/s
occurs at 64 processes, with a corresponding
network data rate of 121 MiB/s
Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
network access by measuring the rate at which we can download the chunks
into RAM (without decompressing) using the same parallel and asynchronous
architecture. Here, the maximum rate of 1.2GiB/s for a c5.9xlarge instance
and 2.4GiB/s for a c5n.9xlarge (high performance network) instance
was achieved with 32 processes. 
% Richard Durbin asked whether we were just using up all of a scarce 
% resource on this point.
While these measurments are simplistic and may not be an 
accurate predictor of sustained bandwidth, it is still clear that 
genotype data can be fetched
from S3 around an order of magnitude more quickly 
than it can be decompressed into RAM. 
We have also not measured
the achievable bandwidth while scaling out across many instances,
but given the ubiquity of cloud-based services and the scale 
at which they operate, it seems reasonable to assume that the claimed
terabits per second of aggregate bandwidth from
S3~\cite{s3-performance-optimising} are achievable.

@@

\subsection{AWS parallel Zarr prototype}
This section describes the methodology used for the AWS parallel Zarr
throughput measurements in Fig~\ref{fig-s3-throughput}
and Fig~\ref{fig-s3-network-throughput}, and timing measurements
detailed in the cloud computing results.
These measurements were run on EC2 instances using jupyter notebooks
hosted on Genomics England's Lifebit CloudOS platform. VCF Zarr chunks
were hosted in a standard profile S3 bucket. Both instance and bucket were
in the same availability zone in the eu-west-2 region. Each
of the parallel Python processes used the aioboto3 library (version 13.2.0) to
fetch chunks concurrently from S3. The underlying aiohttp library
limits the concurrent fetching to 100 requests per process. For decompressing
and computing af-dist the same architecture was used but as fetched chunks
became available in memory in each process, they were decompressed and
computed sequentially. Decompression was performed by the blosc library
without the use of zarr.

@@

\section{Acknowledgements}
We would like to thank Gil McVean and Ben Neale for helpful discussions and
comments on the manuscript.
This research was made possible through access to data in the National Genomic
Research Library, which is managed by Genomics England Limited (a wholly owned
company of the Department of Health and Social Care). The National Genomic
Research Library holds data provided by patients and collected by the NHS as
part of their care and data collected as part of their participation in
research. The National Genomic Research Library is funded by the National
Institute for Health Research and NHS England. The Wellcome Trust, Cancer
Research UK and the Medical Research Council have also funded research
infrastructure.

>>>>

\subsection{Future applications: Cloud computing}
To explore the use of cloud computing services with VCF Zarr,
Genomics England provided access to Aggv2 data 
in VCF Zarr format within their secure
cloud environment. The VCF Zarr files for Chromosome 2
described in Table~\ref{tab-genomics-england-data}
were copied to a standard AWS S3 bucket.
Using this data we repeated the benchmarks above on the same subset
of 562,640 variants and 78,195 samples using the same
single-threaded sequential Zarr code.
Extracting the position array required 15s CPU, 23.8s elapsed.
The af-dist calculation required 2.70min CPU, 5.23min elapsed.
Creating the filter mask required 3.05min CPU, 11.15min elapsed.

While these benchmarks are comparable with results obtained on the
shared file-system in the Genomics England research environment,
the sequential single-threaded access patterns used do not map
well to modern cloud computing environments.
As each request to a cloud object store has a high latency, it
is important to have many requests running concurrently to achieve
good performance~\citep{durner2023exploiting}.
To explore the limits of performance we developed a simple prototype
Zarr implementation, designed to maximise concurrency using asynchronous
IO and multiple processes (see Methods for details).
Fig~\ref{fig-s3-throughput} shows the rate at which genotypes
can be decoded to RAM as well as rate at which we can compute
af-dist (using Numba to classify genotypes, as before)
as we increase the number of parallel processes on
an EC2 c5.24xlarge instance with 96 vCPUs (48 physical).
The maximum decompression rate of 47.4GiB/s
occurs at 64 processes (each with 100 concurrent requests), 
with a corresponding network data rate of 121 MiB/s
Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
network access by measuring the rate at which we can download the chunks
into RAM (without decompressing) using the same parallel and asynchronous
architecture. Here, the maximum rate of 1.2GiB/s for a c5.9xlarge instance
and 2.4GiB/s for a c5n.9xlarge (high performance network) instance
was achieved with 32 processes. 
% Richard Durbin asked whether we were just using up all of a scarce 
% resource on this point.
While these measurments are simplistic and may not be an 
accurate predictor of sustained bandwidth, it is still clear that 
genotype data can be fetched
from S3 around an order of magnitude more quickly 
than it can be decompressed into RAM. 
We have also not measured
the achievable bandwidth while scaling out across many instances,
but given the ubiquity of cloud-based services and the scale 
at which they operate, it seems reasonable to assume that the claimed
terabits per second of aggregate bandwidth from
S3~\cite{s3-performance-optimising} are achievable.

@@

\subsection{AWS parallel Zarr prototype}
This section describes the methodology used for the AWS parallel Zarr
throughput measurements in Fig~\ref{fig-s3-throughput}
and Fig~\ref{fig-s3-network-throughput}, and timing measurements
detailed in the cloud computing results.
These measurements were run on EC2 instances using Jupyter notebooks
hosted on Genomics England's Lifebit CloudOS platform. VCF Zarr chunks
were hosted in a standard profile S3 bucket. Both instance and bucket were
in the same availability zone in the eu-west-2 region. Each
of the parallel Python processes used the aioboto3 library (version 13.2.0) to
fetch chunks concurrently from S3.
% The underlying aiohttp library
% limits the concurrent fetching to 100 requests per process. 
For decompressing
and computing af-dist the same architecture was used but as fetched chunks
became available in memory in each process, they were decompressed and
processed syncronously. Decompression was performed by using the 
numcodecs library~\citep{numcodecs}.

This prototype did not use any existing Zarr implementation as 
the levels of concurrency involved require specialised 
asynchronous programming techniques that are not widely supported.
The fact that such a high-performance prototype can be implemented
in a notebook with no Zarr-specific library support is a testament to the 
transparent simplicity of the Zarr format.

@@

\section{Acknowledgements}
We are grateful to Augusto Rendon and the Genomics 
England support team for facilitating our cloud computing experiments,
and to Gil McVean and Ben Neale for helpful discussions and comments on the 
manuscript.

This research was made possible through access to data in the National Genomic
Research Library, which is managed by Genomics England Limited (a wholly owned
company of the Department of Health and Social Care). The National Genomic
Research Library holds data provided by patients and collected by the NHS as
part of their care and data collected as part of their participation in
research. The National Genomic Research Library is funded by the National
Institute for Health Research and NHS England. The Wellcome Trust, Cancer
Research UK and the Medical Research Council have also funded research
infrastructure.
